---
title: "Advent of Code: Day 3"
author: "Caroline Falvey"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(stringr)
setwd("/Users/carolinefalvey/Desktop/Duke")

```

### Day 3: Rucksack Reorganization

## Part I:

One Elf has the important job of loading all of the rucksacks with
supplies for the jungle journey. Unfortunately, that Elf didn't quite
follow the packing instructions, and so a few items now need to be
rearranged.

Each rucksack has two large compartments. All items of a given type are
meant to go into exactly one of the two compartments. The Elf that did
the packing failed to follow this rule for exactly one item type per
rucksack.

The Elves have made a list of all of the items currently in each
rucksack (your puzzle input), but they need your help finding the
errors. Every item type is identified by a single lowercase or uppercase
letter (that is, a and A refer to different types of items).

The list of items for each rucksack is given as characters all on a
single line. A given rucksack always has the same number of items in
each of its two compartments, so the first half of the characters
represent items in the first compartment, while the second half of the
characters represent items in the second compartment.

For example, suppose you have the following list of contents from six
rucksacks:

<br>

|                                                                    |
|--------------------------------------------------------------------|
| vJrwpWtwJgWrhcsFMMfFFhFp jqHRNqRjqzjGDLGLrsFMfFZSrLrFZsSL          |
| PmmdzqPrVvPwwTWBwg wMqvLMZHhHMvwLHjbvcjnnSBnvTQFn ttgJtRGJQctTZtZT |
| CrZsJsPPZsGzwwsLwLmpwMDw                                           |

<br>

The first rucksack contains the items vJrwpWtwJgWrhcsFMMfFFhFp, which
means its first compartment contains the items vJrwpWtwJgWr, while the
second compartment contains the items hcsFMMfFFhFp. The only item type
that appears in both compartments is lowercase p.

The second rucksack's compartments contain jqHRNqRjqzjGDLGL and
rsFMfFZSrLrFZsSL. The only item type that appears in both compartments
is uppercase L.

The third rucksack's compartments contain PmmdzqPrV and vPwwTWBwg; the
only common item type is uppercase P.

The fourth rucksack's compartments only share item type v. The fifth
rucksack's compartments only share item type t. The sixth rucksack's
compartments only share item type s.

To help prioritize item rearrangement, every item type can be converted
to a priority:

-   Lowercase item types a through z have priorities 1 through 26.

-   Uppercase item types A through Z have priorities 27 through 52.

In the above example, the priority of the item type that appears in both
compartments of each rucksack is 16 (p), 38 (L), 42 (P), 22 (v), 20 (t),
and 19 (s); the sum of these is 157.

**Find the item type that appears in both compartments of each rucksack.
What is the sum of the priorities of those item types?**

<br>

```{r}
# Read in data from website:
data <- read_csv("compartments.txt", col_names = F)

# Convert data to vector & initialize matrix for split data
data <- as.vector(unlist(data))
split <- matrix(NA, nrow = length(data), ncol = 2)

# Split strings in half to create 1st and 2nd compartments
for (i in seq_along(data)) {
  split[i, 1] <- substr(data[i], 1, nchar(data[i]) / 2)
  split[i, 2] <- substr(data[i], nchar(data[i]) / 2 + 1, nchar(data[i]))
}

# View original and split data
view(cbind.data.frame(data, split))

# Initialize vector to store repeated letters
letter <- rep(NA, nrow(split))


for (j in 1:nrow(split)) {
  # Split strings into separate letters
  test1 <- unlist(strsplit(split[j, 1], split = ""))
  test2 <- unlist(strsplit(split[j, 2], split = ""))

  # Confirm which letter is repeated in each string
  key <- test1 %in% test2 
  
  # Store repeated letter into letter vector
  letter[j] <- test1[which(key)]
}

# Combine original data, split data, and repeated letter
full_set <- cbind.data.frame(data, split, letter)


# Combine lower case and upper case letter sets A-Z
ids <- c(letters, LETTERS)

for (k in 1:nrow(full_set)) {
  # Calculate priority for each compartment based off repeated letter
  full_set$priority[k] <- which(full_set$letter[k] == ids)
}


# Sum priorities
sum(full_set$priority) ## Answer 7428


```

<br>

Your puzzle answer was 7428.

<br>

## Part II:

As you finish identifying the misplaced items, the Elves come to you
with another issue.

For safety, the Elves are divided into groups of three. Every Elf
carries a badge that identifies their group. For efficiency, within each
group of three Elves, the badge is the only item type carried by all
three Elves. That is, if a group's badge is item type B, then all three
Elves will have item type B somewhere in their rucksack, and at most two
of the Elves will be carrying any other item type.

The problem is that someone forgot to put this year's updated
authenticity sticker on the badges. All of the badges need to be pulled
out of the rucksacks so the new authenticity stickers can be attached.

Additionally, nobody wrote down which item type corresponds to each
group's badges. The only way to tell which item type is the right one is
by finding the one item type that is common between all three Elves in
each group.

Every set of three lines in your list corresponds to a single group, but
each group can have a different badge item type. So, in the above
example, the first group's rucksacks are the first three lines:

<br>

|                                   |
|-----------------------------------|
| vJrwpWtwJgWrhcsFMMfFFhFp          |
| jqHRNqRjqzjGDLGLrsFMfFZSrLrFZsSL  |
| PmmdzqPrVvPwwTWBwg                |

<br>

And the second group's rucksacks are the next three lines:

<br>

|                                |
|--------------------------------|
| wMqvLMZHhHMvwLHjbvcjnnSBnvTQFn |
| ttgJtRGJQctTZtZT               |
| CrZsJsPPZsGzwwsLwLmpwMDw       |

<br>

In the first group, the only item type that appears in all three
rucksacks is lowercase r; this must be their badges. In the second
group, their badge item type must be Z.

Priorities for these items must still be found to organize the sticker
attachment efforts: here, they are 18 (r) for the first group and 52 (Z)
for the second group. The sum of these is 70.

**Find the item type that corresponds to the badges of each three-Elf
group. What is the sum of the priorities of those item types?**

<br>

```{r}

# Convert original data into a 3x100 matrix
matrix <- matrix(data, nrow = 3, ncol = length(data) / 3, byrow = F)
# Initialize vector to store new repeated letters
new_letters <- rep(NA, ncol(matrix))


for (n in 1:ncol(matrix)) {
  # Split strings into separate letters
  set1 <- unlist(strsplit(matrix[1, n], split = ""))
  set2 <- unlist(strsplit(matrix[2, n], split = ""))
  set3 <- unlist(strsplit(matrix[3, n], split = ""))

  # Confirm which letters are repeated in 1st string and 2nd string
  key1 <- unique(set1[which(set1 %in% set2)])
  # Confirm which letters from above are repeated in 3rd string
  key2 <- key1[which(key1 %in% set3)]
  # Store repeated letter in new letter vector
  new_letters[n] <- key2
}

# Initialize new priority vector
new_priority <- rep(NA, length(new_letters))


for (m in seq_along(new_letters)) {
  # Calculate priority for each compartment based off repeated letter
  new_priority[m] <- which(new_letters[m] == ids)
}


# Sum priorities
sum(new_priority) ## Answer: 2650

```


<br>

Your puzzle answer was 2650.
