---
title: 'Advent of Code: Day 1'
author: "Caroline Falvey"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

### Day 4: Camp Cleanup

## Part I:

Space needs to be cleared before the last supplies can be unloaded from
the ships, and so several Elves have been assigned the job of cleaning
up sections of the camp. Every section has a unique ID number, and each
Elf is assigned a range of section IDs.

However, as some of the Elves compare their section assignments with
each other, they've noticed that many of the assignments overlap. To try
to quickly find overlaps and reduce duplicated effort, the Elves pair up
and make a big list of the section assignments for each pair (your
puzzle input).

For example, consider the following list of section assignment pairs:

<br>

2-4,6-8

2-3,4-5

5-7,7-9

2-8,3-7

6-6,4-6

2-6,4-8

<br>

For the first few pairs, this list means:

-   Within the first pair of Elves, the first Elf was assigned sections
    2-4 (sections 2, 3, and 4), while the second Elf was assigned
    sections 6-8 (sections 6, 7, 8).

-   The Elves in the second pair were each assigned two sections.

-   The Elves in the third pair were each assigned three sections: one
    got sections 5, 6, and 7, while the other also got 7, plus 8 and 9.

This example list uses single-digit section IDs to make it easier to
draw; your actual list might contain larger numbers. Visually, these
pairs of section assignments look like this:

<br>

.234..... 2-4

.....678. 6-8

.23...... 2-3

...45.... 4-5

....567.. 5-7

......789 7-9

.2345678. 2-8

..34567.. 3-7

.....6... 6-6

...456... 4-6

.23456... 2-6

...45678. 4-8

<br>

Some of the pairs have noticed that one of their assignments fully
contains the other. For example, 2-8 fully contains 3-7, and 6-6 is
fully contained by 4-6. In pairs where one assignment fully contains the
other, one Elf in the pair would be exclusively cleaning sections their
partner will already be cleaning, so these seem like the most in need of
reconsideration. In this example, there are 2 such pairs.

**In how many assignment pairs does one range fully contain the other?**

<br>

```{r}

# Read in raw data
sections <- read_csv("sections.txt", col_names = F)
# Rename the columns for clarity
colnames(sections) <- c("elf1", "elf2")


# Split the assignments by the dash to get the first and last
# section assignment for each elf
elf1_split <- strsplit(sections$elf1, "-")
elf2_split <- strsplit(sections$elf2, "-")

# Create new variables for first & last sections for elf 1
elf1_low <- as.numeric(unlist(lapply(elf1_split, `[[`, 1)))
elf1_upp <- as.numeric(unlist(lapply(elf1_split, `[[`, 2)))

# Create new variable for first & last sections for elf 2
elf2_low <- as.numeric(unlist(lapply(elf2_split, `[[`, 1)))
elf2_upp <- as.numeric(unlist(lapply(elf2_split, `[[`, 2)))


# Create a new data frame with the initial assignments &
# the upper and lower sections for both elves
sections_clean <- cbind.data.frame(
  sections,
  elf1_low, elf1_upp,
  elf2_low, elf2_upp
)

# Initialize a result vector
result_vec <- rep(NA, nrow(sections_clean))



for (i in 1:nrow(sections_clean)) {
  if (sections_clean$elf1_low[i] <= sections_clean$elf2_low[i] &&
    sections_clean$elf1_upp[i] >= sections_clean$elf2_upp[i]) {
    # Determine if the 2nd elf's sections are contained in the 1st elf's sections
    result_vec[i] <- "2 in 1"
  } else {
    if (sections_clean$elf2_low[i] <= sections_clean$elf1_low[i] &&
      sections_clean$elf2_upp[i] >= sections_clean$elf1_upp[i]) {
      # Determine if the 1st elf's sections are contained in the 2nd elf's sections
      result_vec[i] <- "1 in 2"
    } else {
      # If neither of the above, indicate that the assignment was done correctly
      result_vec[i] <- "correct"
    }
  }
}


# Subtract the number of correct assignments from the total number of
# assignments to get the total number of incorrect assignments
length(result_vec) - sum(result_vec == "correct")
## Answer 462

```

<br>

Your puzzle answer was `r {length(result_vec) - sum(result_vec == "correct")}`.

## Part II:

It seems like there is still quite a bit of duplicate work planned.
Instead, the Elves would like to know the number of pairs that overlap
at all.

In the above example, the first two pairs (`2-4,6-8` and `2-3,4-5`)
don't overlap, while the remaining four pairs (`5-7,7-9`, `2-8,3-7`,
`6-6,4-6`, and `2-6,4-8`) do overlap:

-   `5-7,7-9` overlaps in a single section, 7.

-   `2-8,3-7` overlaps all of the sections 3 through 7.

-   `6-6,4-6` overlaps in a single section, 6.

-   `2-6,4-8` overlaps in sections 4, 5, and 6.

So, in this example, the number of overlapping assignment pairs is 4.

**In how many assignment pairs do the ranges overlap?**

<br>

```{r}

# Initialize a second result vector
result_vec_2 <- rep(NA, nrow(sections_clean))

for (j in 1:nrow(sections_clean)) {
  # Create a vector of all sections sections assigned to elf 1
  elf1_total <- elf1_low[j]:elf1_upp[j]
  # Create a vector of all sections assigned to elf 2
  elf2_total <- elf2_low[j]:elf2_upp[j]

  # Determine if there are any overlapping sections
  duplicates <- sum(elf1_total %in% elf2_total)

  # If there are any duplicates, result = T
  # If there are not duplicates, result = F
  result_vec_2[j] <- ifelse(duplicates > 0, T, F)
}

# Sum the result vector to calculate the number of assignments w duplicates
sum(result_vec_2)
## Answer: 835

```

<br>

Your puzzle answer was `r {sum(result_vec_2)}`.
