---
title: 'Advent of Code: Day 1'
author: "Caroline Falvey"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

### Day 2: Rock Paper Scissors

## Part I:

The Elves begin to set up camp on the beach. To decide whose tent gets
to be closest to the snack storage, a giant Rock Paper Scissors
tournament is already in progress.

Rock Paper Scissors is a game between two players. Each game contains
many rounds; in each round, the players each simultaneously choose one
of Rock, Paper, or Scissors using a hand shape. Then, a winner for that
round is selected: Rock defeats Scissors, Scissors defeats Paper, and
Paper defeats Rock. If both players choose the same shape, the round
instead ends in a draw.

Appreciative of your help yesterday, one Elf gives you an encrypted
strategy guide (your puzzle input) that they say will be sure to help
you win. "The first column is what your opponent is going to play: A for
Rock, B for Paper, and C for Scissors. The second column--" Suddenly,
the Elf is called away to help with someone's tent.

The second column, you reason, must be what you should play in response:
X for Rock, Y for Paper, and Z for Scissors. Winning every time would be
suspicious, so the responses must have been carefully chosen.

The winner of the whole tournament is the player with the highest score.
Your total score is the sum of your scores for each round. The score for
a single round is the score for the shape you selected (1 for Rock, 2
for Paper, and 3 for Scissors) plus the score for the outcome of the
round (0 if you lost, 3 if the round was a draw, and 6 if you won).

Since you can't be sure if the Elf is trying to help you or trick you,
you should calculate the score you would get if you were to follow the
strategy guide.

For example, suppose you were given the following strategy guide:

|     |     |
|:---:|:---:|
|  A  |  Y  |
|  B  |  X  |
|  C  |  Z  |

This strategy guide predicts and recommends the following:

-   In the first round, your opponent will choose Rock (A), and you
    should choose Paper (Y). This ends in a win for you with a score of
    8 (2 because you chose Paper + 6 because you won).

-   In the second round, your opponent will choose Paper (B), and you
    should choose Rock (X). This ends in a loss for you with a score of
    1 (1 + 0).

-   The third round is a draw with both players choosing Scissors,
    giving you a score of 3 + 3 = 6. In this example, if you were to
    follow the strategy guide, you would get a total score of 15 (8 +
    1 + 6).

**What would your total score be if everything goes exactly according to
your strategy guide?**

<br>

```{r}

# Read in raw data
strategy <- read_csv("rps_2.txt", col_names = F)

# Convert the raw data from a data frame to a vector
strategy_2 <- as.vector(unlist(strategy))
# Split the two characters from the raw data into separate columns
split <- strsplit(strategy_2, " ")

# Designate the first row as the opponent's choice
opp <- unlist(lapply(split, `[[`, 1))
# Designate the second row as our choice
me <- unlist(lapply(split, `[[`, 2))

# Create a data frame with the opponent's choice vs our choice
strategy_clean <- cbind.data.frame(opp, me)

# Rename the opponent's move so it matches X, Y, Z format
# A = X (rock), B = Y (paper), C = Z (scissors)
strategy_clean$opp[strategy_clean$opp == "A"] <- "X"
strategy_clean$opp[strategy_clean$opp == "B"] <- "Y"
strategy_clean$opp[strategy_clean$opp == "C"] <- "Z"

# Designate points based on our choice in the round
# rock = 1 pt, paper = 2 pts, scissors = 3 pts
strategy_clean$initial_pts <- ifelse(strategy_clean$me == "X", 1,
  ifelse(strategy_clean$me == "Y", 2, 3)
)



# Determine if we win, lose, or draw vs opponent for all games played
for (i in 1:nrow(strategy_clean)) {
  if (strategy_clean$opp[i] == strategy_clean$me[i]) {
    strategy_clean$outcome[i] <- 3 # draw game (3 pts)
  } else {
    if (strategy_clean$opp[i] == "X" && strategy_clean$me[i] == "Z") {
      strategy_clean$outcome[i] <- 0 # rock beats scissors (0 pts)
    } else {
      if (strategy_clean$opp[i] == "Y" && strategy_clean$me[i] == "X") {
        strategy_clean$outcome[i] <- 0 # paper beats rock (0 pts)
      } else {
        if (strategy_clean$opp[i] == "Z" && strategy_clean$me[i] == "Y") {
          strategy_clean$outcome[i] <- 0 # scissors beat paper (0 pts)
        } else {
          strategy_clean$outcome[i] <- 6 # all other options are a win (6 pts)
        }
      }
    }
  }
}

# Calculate the total points from each round (i.e. initial pts + match outcome)
strategy_clean <- strategy_clean |>
  mutate(total = initial_pts + outcome)

# Sum the total points
sum(strategy_clean$total) 
## Answer = 10941

```

<br>

Your puzzle answer was `r {sum(strategy_clean$total)}`.

## Part II:

The Elf finishes helping with the tent and sneaks back over to you.
"Anyway, the second column says how the round needs to end: `X` means
you need to lose, `Y` means you need to end the round in a draw, and `Z`
means you need to win. Good luck!"

The total score is still calculated in the same way, but now you need to
figure out what shape to choose so the round ends as indicated. The
example above now goes like this:

-   In the first round, your opponent will choose Rock (`A`), and you
    need the round to end in a draw (`Y`), so you also choose Rock. This
    gives you a score of 1 + 3 = 4.

-   In the second round, your opponent will choose Paper (`B`), and you
    choose Rock so you lose (`X`) with a score of 1 + 0 = 1.

-   In the third round, you will defeat your opponent's Scissors with
    Rock for a score of 1 + 6 = 7.

Now that you're correctly decrypting the ultra top secret strategy
guide, you would get a total score of 12.

**Following the Elf's instructions for the second column, what would
your total score be if everything goes exactly according to your
strategy guide?**

<br>

```{r}
## Second Column:
# X means we need to lose,
# Y means we need to end the round in a draw, and
# Z means we need to win

# Create new data frame for the new rules
strategy_new <- strategy_clean |>
  select(opp, me)

# X beats Z
# Y beats X
# Z beats Y

for (j in 1:nrow(strategy_new)) {
  if (strategy_new$me[j] == "Y") { 
    # Game must end in draw. Our move will be the same as opponent's move
    strategy_new$turn[j] <- strategy_new$opp[j]
    # Draw game (3 pts)
    strategy_new$match_pt[j] <- 3 
  } else {
    if (strategy_new$me[j] == "X" && strategy_new$opp[j] == "X") {
      # Game must end in L. Opponent chooses rock, we choose scissors
      strategy_new$turn[j] <- "Z"
      # Loss (0 pts)
      strategy_new$match_pt[j] <- 0
    } else {
      if (strategy_new$me[j] == "X" && strategy_new$opp[j] == "Y") {
        # Game must end in L. Opponent chooses paper, we choose rock
        strategy_new$turn[j] <- "X"
        # Loss (0 pts)
        strategy_new$match_pt[j] <- 0
      } else {
        if (strategy_new$me[j] == "X" && strategy_new$opp[j] == "Z") {
          # Game must end in L. Opponent chooses scissors, we choose paper
          strategy_new$turn[j] <- "Y"
          # Loss (0 pts)
          strategy_new$match_pt[j] <- 0
        } else {
          if (strategy_new$me[j] == "Z" && strategy_new$opp[j] == "X") {
            # Game must end in W. Opponent chooses rock, we choose paper
            strategy_new$turn[j] <- "Y"
            # Win (6 pts)
            strategy_new$match_pt[j] <- 6
          } else {
            if (strategy_new$me[j] == "Z" && strategy_new$opp[j] == "Y") {
              # Game must end in W. Opponent chooses paper, we choose scissors
              strategy_new$turn[j] <- "Z"
              # Win (6 pts)
              strategy_new$match_pt[j] <- 6
            } else {
              # Game must end in W. Opponent chooses scissors, we choose rock
              strategy_new$turn[j] <- "X"
              # Win (6 pts)
              strategy_new$match_pt[j] <- 6
            }
          }
        }
      }
    }
  }
}

# Check data frame to make sure data is printing as expected
head(strategy_new)

# Designate points based on our choice in the round
# rock = 1 pt, paper = 2 pts, scissors = 3 pts
strategy_new$turn_pts <- ifelse(strategy_new$turn == "X", 1,
  ifelse(strategy_new$turn == "Y", 2, 3)
)

# Calculate the total points from each round (i.e. initial pts + match outcome)
strategy_new <- strategy_new |>
  mutate(total_pts = match_pt + turn_pts)


# Sum the total points
sum(strategy_new$total_pts) 
## Answer = 13071

```

<br>

Your puzzle answer was `r {sum(strategy_new$total_pts)}`.
